---
title: "Other useful visualizations"
output: 
  html_notebook: 
    fig_width: 8
    fig_height: 6
editor_options:
  markdown:
    wrap: 72
---

# 2. Other useful visualizations

In this section, we will practice how to visualize with treemap and geographical maps,
which are both commonly used in economic complexity analysis. Install and load necessary packages:

```{r}
package_list <-
  c(
    "tidyverse",
    "viridis",
    "sf",
    "treemap",
    "rnaturalearth",
    "rnaturalearthdata"
  )
# Check if packages are already installed, otherwise install them
to_install <-
  package_list[!(package_list %in% installed.packages()[, "Package"])]
if (length(to_install)) {
  install.packages(to_install)
}
library(tidyverse)
library(treemap)
library(viridis)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

Let's load the previously used output from the atlas of economic complexity for 2015.

```{r}
if (!file.exists("df_ec.tsv")) {
  download.file("https://www.dropbox.com/s/510nkpc7mwnjvbp/df_ec.tsv?dl=1", "df_ec.tsv")
}
df_ec <- read_tsv("df_ec.tsv")
head(df_ec)
```

## Treemap

Again, let's use Japan as an example, aggregate to 2 digits level:
```{r}
df_jpn <- df_ec %>%
  filter(location_code == "JPN", export_value > 0) %>%
  select(hs_product_code, export_value) %>%
  mutate(
    category = if_else(str_length(hs_product_code) == 4, substring(hs_product_code, 1, 1), "service"),
    hs_product_code = if_else(str_length(hs_product_code) == 4, substring(hs_product_code, 1, 2), hs_product_code)
  ) %>%
  group_by(hs_product_code, category) %>%
  summarise(export_value = sum(export_value))
head(df_jpn)
```

Load product names and merge:
```{r}
if (!file.exists("hs_product.zip")) {
  download.file("http://intl-atlas-downloads.s3.amazonaws.com/17.0/hs_product.zip", "hs_product.zip")
}
hsdf <- read_csv("hs_product.zip", col_types = "c") %>% select(hs_product_code, hs_product_name_short_en)
df_jpn <- df_jpn %>%
  left_join(hsdf) %>%
  left_join(hsdf %>% rename(category = hs_product_code, categoryname = hs_product_name_short_en)) %>%
  mutate(categoryname = if_else(is.na(categoryname), category, categoryname))
head(df_jpn)
```


Use the `treemap` package to plot a nested treemap

```{r}
treemap(df_jpn,
  index = c("categoryname", "hs_product_name_short_en"),
  vSize = "export_value",
  type = "index",
  title = "Export basket of Japan in 2015"
)
```

## Geographical map

The R geographcial packages have been moving to a `sf` package based ecosystem, where the spatial
dataset becomes a dataframe-like structure, and the manipulation and visualization could be handled
easily within the tidyverse framework. Load the worldmap and check the data:

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
head(world)
```
Plotting is simply with `ggplot2`:

```{r}
ggplot(data = world) +
  geom_sf()
```
Manipulate the data as a dataframe:

```{r}
world <- world %>%
  filter(pop_est > 0, continent != "Antarctica") %>%
  mutate(gdp_per_cap = gdp_md_est / pop_est)
```
The data comes with GDP information, let's use it to color the countries and add additional elements

```{r}
ggplot(data = world) +
  geom_sf(aes(fill = log(gdp_per_cap))) +
  scale_fill_viridis() +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("World GDP per Capita") +
  theme_bw()
```
Let's extract the ECI of each country

```{r}
df_eci <- df_ec %>%
  select(location_code, location_name_short_en, eci) %>%
  distinct() %>%
  drop_na()
head(df_eci)
```

Join the data, and plot:
```{r}
worldeci <- world %>%
  left_join(df_eci, by = c("adm0_a3" = "location_code"))
ggplot(data = worldeci) +
  geom_sf(aes(fill = eci)) +
  scale_fill_viridis() +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Country economic complexity index") +
  theme_bw()
```

